Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-20 paper.tex
More citations for forks of ms

The abilily to efficiently process simulation results is
particularly important in simulation-based inference methods
such as Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and machine learning based
approaches~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable,
sanchez2020deep}. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence
data, and methods avoid the bottleneck of parsing
text-based file formats to compute these statistics
by either developing their own
simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
or creating forked versions of existing
simulators~\cite[e.g.][]{pavlidis2010msabc,quinto2018modeling},
tightly integrated with the inference method.
While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error (and propagation of bugs).
Modern approaches to ABC such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} use large
numbers of weakly informative statistics,
making the need to efficiently compute statistics from simulation
results even more acute.

Classical text based output formats like \ms\ are inefficient to process,
but also lacking a great deal of important information.
The tree-by-tree topology information output by simulators in Newick
format lacks any concept of node (in the tree sequence sense) identity,
and means that we cannot reliably infer information about ancestors
from the output \jkcomment{Also there's the inherent limitation of
newick using branch lenghts; I think there's some citations for
the issues this causes. Probably in the original msprime paper.}.
This has been worked around by using approximate definitions of
concepts like identity by descent (cite carmi paper) \jkcomment{
any other computation methods developed just to work around the
crappiness of ms output?}
Numerous forks
of simulators have been created to access information not provided
in the output. For example, \ms\  has been forked to
output information about migrating segments~\citep{rosenzweig2016powerful},
information about ancestral lineages~\citep{chen2013asymptotic},
% Any more?
and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
has in turn been forked to output information on local
ancestry~\citep{racimo2017archaic}.
All of this information
is either directly available by default in \msprime, or can be optionally
stored via options such as \texttt{record\_migrations} or
\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
can be efficiently and conveniently processed via \tskit\ APIs.

>>>>

The abilily to efficiently process simulation results is
particularly important in simulation-based inference methods
such as Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and machine learning based
approaches~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable,
sanchez2020deep}. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence
data, and methods avoid the bottleneck of parsing
text-based file formats to compute these statistics
by either developing their own
simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
or creating forked versions of existing
simulators~\cite[e.g.][]{thornton2006approximate,
hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
tightly integrated with the inference method.
While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error (and propagation of bugs).
Modern approaches to ABC such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} use large
numbers of weakly informative statistics,
making the need to efficiently compute statistics from simulation
results even more acute.

Classical text based output formats like \ms\ are inefficient to process,
but also lacking a great deal of important information.
The tree-by-tree topology information output by simulators in Newick
format lacks any concept of node (in the tree sequence sense) identity,
and means that we cannot reliably infer information about ancestors
from the output \jkcomment{Also there's the inherent limitation of
newick using branch lenghts; I think there's some citations for
the issues this causes. Probably in the original msprime paper.}.
This has been worked around by using approximate definitions of
concepts like identity by descent (cite carmi paper) \jkcomment{
any other computation methods developed just to work around the
crappiness of ms output?}
Numerous forks
of simulators have been created to access information not provided
in the output. For example, \ms\  has been forked to
output information about migrating segments~\citep{rosenzweig2016powerful},
information about ancestral lineages~\citep{chen2013asymptotic},
an array of population genetic statistics~\citep{ramos2007mlcoalsim}
% Any more?
and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
has in turn been forked to output information on local
ancestry~\citep{racimo2017archaic}.
All of this information
is either directly available by default in \msprime, or can be optionally
stored via options such as \texttt{record\_migrations} or
\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
can be efficiently and conveniently processed via \tskit\ APIs.
